---
title: "Relative Risk associated with Education"
author: "Richard Martin"
format:
  html:
    page-layout: full
    code-fold: true
    code-summary: "Show the code"
---
```{r, include=FALSE}
library(tidyverse)
library(here)
library(readxl)
library(plotly)
library(conflicted)
conflicts_prefer(dplyr::filter)
bad_occ <- read_csv(here("out","bad_occ.csv"))
relative_risk <- read_csv(here("out","relative_risk.csv"))
cut_off=.2
my_dt <- function(tbbl, length) {
  DT::datatable(tbbl,
    extensions = "Buttons",
    rownames = FALSE,
    options = list(
      columnDefs = list(list(className = "dt-center", targets = "_all")),
      paging = TRUE,
      scrollX = TRUE,
      scrollY = TRUE,
      searching = TRUE,
      ordering = TRUE,
      dom = "Btip",
      buttons = list(
        list(extend = "csv", filename = "some_file_name"),
        list(extend = "excel", filename = "some_file_name")
      ),
      pageLength = length
    )
  )
}
```

```{r, eval=FALSE}
#constants------------
cut_off=.2
#functions---------------
my_dt <- function(tbbl, length) {
  DT::datatable(tbbl,
    extensions = "Buttons",
    rownames = FALSE,
    options = list(
      columnDefs = list(list(className = "dt-center", targets = "_all")),
      paging = TRUE,
      scrollX = TRUE,
      scrollY = TRUE,
      searching = TRUE,
      ordering = TRUE,
      dom = "Btip",
      buttons = list(
        list(extend = "csv", filename = "some_file_name"),
        list(extend = "excel", filename = "some_file_name")
      ),
      pageLength = length
    )
  )
}
get_relative_risk <- function(edu, tbbl){
  prob_bad_conditional_on_education <- tbbl|>
    filter(Education==edu)|>
    group_by(occ_type)|>
    summarize(value=sum(value))|>
    ungroup()|>
    mutate(value=value/sum(value))|>
    filter(occ_type=="bad")|>
    pull(value)

  prob_bad_conditional_on_not_education <- tbbl|>
    filter(Education!=edu)|>
    group_by(occ_type)|>
    summarize(value=sum(value))|>
    ungroup()|>
    mutate(value=value/sum(value))|>
    filter(occ_type=="bad")|>
    pull(value)

  relative_risk <- prob_bad_conditional_on_education/prob_bad_conditional_on_not_education
}
#the code--------------------
cip_noc_all <- read_csv(here("out","cip_noc_all_ages_processed.csv"))|>
  rename(NOCtemp=NOC)|>
  mutate(NOC=str_sub(NOCtemp, 1,5),
         Description=str_sub(NOCtemp, 7), .before=everything())|>
  select(-NOCtemp)

median_salaries <- read_excel(here("data","median_salary.xlsx"))|>
  select(NOC, median_salary=contains("calculated"))

paths <- read_excel(here("out","top_paths_to_and_from.xlsx"), sheet = "Paths from Education all ages")|>
  select(highest=`For students who attained`,CIP=`in this field of study`, paths=`# of educations with share > 1%`)|>
  distinct()

skills <- read_excel(here("data", "skills_original.xlsx"))|>
  mutate(NOC=str_pad(noc2021, 5, "left", "0"), .before=everything())|>
  select(-noc2021)|>
  group_by(NOC)|>
  summarize(average_skill=round(mean(value_mean),2))

low_median_salary <- median_salaries|>
  slice_min(median_salary, prop = cut_off)

low_skills <- skills|>
  slice_min(average_skill, prop = cut_off)

bad_occ <- inner_join(low_median_salary, low_skills)|>
  mutate(occ_type="bad")|>
  select(NOC, occ_type)

joined <- full_join(cip_noc_all, bad_occ)|>
  mutate(occ_type=if_else(is.na(occ_type), "good",  occ_type))

nested <- tibble(Education=unique(joined$Education), data=list(joined))|>
  na.omit()|>
  mutate(relative_risk=map2_dbl(Education, data, get_relative_risk))

relative_risk <- nested|>
  select(Education, relative_risk)|>
  separate(Education, into=c("CIP","highest"), sep=": ")
```

## Introduction

One important use of the British Columbia Labour Market Outlook is to encourage prospective students to choose fields of study that have a clear/narrow path to high opportunity occupations. Consider optometry, where 91% of people with a degree in optometry work as optometrists. For such narrow paths the LMO can provide *positive* guidance regarding field of study.

However there are many occupations that do not have such a clear linkage with education, which makes education a risky investment. Most people are risk averse^[at least for large gambles, like education], and are willing to sacrifice some expected return in exchange for greater certainty in outcomes.  Here we focus on the relative risk of ending up in a low skill low salary occupation.

We define "bad" occupations to be those that have a median salary in the bottom `r scales::percent(cut_off)`, and an average ONET skill score in the bottom `r scales::percent(cut_off)`, resulting in `r nrow(bad_occ)` "bad" occupations. 

## The "bad" occupations

```{r}
bad_occ|>
  select(-occ_type)|>
  select(NOC, Description, everything())|>
  my_dt(length=-1)
```


## Description of Relative Risk

The relative risk of a given education is defined as:

$$ \frac{prob[bad~occupation~|~this~education]}{prob[bad~occupation~|~all~other~educations]}$$
Note that a relative risk greater/less than 1 indicates that for this field of study you are more/less likely to end up in a "bad" occupation, relative to all other occupations.  If relative risk was exactly 1 then you would be no more or less likely than with any other field of study. 

## The relative risk of ending up in a low salary low skill occupation, by education

```{r, fig.retina=2, fig.height=10, warning=FALSE}
plt <- ggplot(relative_risk, aes(relative_risk, 
                          tidytext::reorder_within(CIP,relative_risk,highest), 
                          text=paste0(
                            "Field of Study: ",
                            CIP, 
                            "\n Relative Risk of low-salary low skill occupation = ",
                            round(relative_risk,2)
                            )))+
  geom_vline(xintercept = 1, colour="grey60")+
  geom_point(alpha=.2)+
  tidytext::scale_y_reordered()+
  scale_x_continuous(trans="log10")+
  facet_wrap(~highest, scales = "free_y")+
  labs(x="Relative Risk",
       y=NULL)+
  theme(axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()) 

ggplotly(plt, tooltip = "text")|>
  config(displayModeBar = FALSE)
```

## The relative risk data:

```{r}
relative_risk|>
  mutate(relative_risk=round(relative_risk, 2))|>
my_dt(length=10)
```



## Conclusion

The LMO *encourages* prospective students to follow educational paths that lead clearly/narrowly to high opportunity occupations.  In contrast, here we are *warning* prospective students about fields of study with a higher risk of ending up in "bad" occupations. This information is critically important to risk averse individuals.

Note that these education investment decisions are also important from a social perspective: as a society we are better off if we can curtail low or negative return education investments.
